\addchap*{Abstract}
\pdfbookmark{Abstract}{abstract}
%
This thesis presents new visualization techniques geared towards two different
applications of engineering simulations: Turbulent combustion and solid
mechanics.
%

%
Direct numerical simulations of turbulent combustion are used as a basis to
develop and validate higher-level combustion models.
%
A focus of interest in the analysis of such simulations is the flame surface,
where most of the chemical reactions take place.
%
In recent years, the computational power of supercomputers has increased much
faster than the performance of storage infrastructures.
%
This has caused the output and storage of simulation data to become the
bottleneck in modern large-scale simulation runs.
%
We introduce two new techniques for the visualization and analysis of the flame
surface in large-scale combustion simulations, before the background of this
storage bottleneck.
%
The first is a space-saving sparse representation for certain types of flames
that allows for the analysis of a larger number of simulation time steps and is
the basis for a new flame visualization technique.
%
The second is an algorithm for tracking the flame surface in-situ during the
simulation in order to circumvent the storage bottleneck and only write to disk
the much smaller results.
%
Both contribute to the continued ability of combustion researchers to analyze
the data produced by their increasingly large simulations.
%

%
Tensor fields belong to some of the most challenging types of data to
visualize, due to their many degrees of freedom.
%
One possibility to break down this complexity is feature-based visualization,
which reduces the data to a set of geometric primitives that represent the
occurrence of some kind of interesting behavior.
%
In scalar and vector fields, a number of line-type features can be extracted
using the parallel vectors operator, which yields locations where two vector
fields are parallel.
%
We translate this operator to tensor fields by introducing the parallel
eigenvectors operator, which yields locations where two tensor fields have
parallel real eigenvectors.
%
We then use this idea to introduce tensor core lines, which mark the centers of
'swirling' behavior of the eigenvectors and are based on vortex core lines in
vector fields.
%
Using this new feature, we can detect twist in the stress tensor fields
produced by solid mechanics simulations.
%