\addchap*{Acknowledgements}
%
Like many others, the document you are holding in your hand or viewing
on your screen is the culmination of several years of hard work and
procrastination, joy and pain, success and disappointment, learning and
teaching, of which only a tiny fraction is represented on these pages.
%
The journey that led to its creation would not have been possible without the
help of some people, and it would not have been as enjoyable, successful, and
educational without several others.
%
I want to express my gratitude to all of them here, for contributing to this
exciting and enriching time of my life.
%

%
Holger, the way you always keep on track and never lose sight of the big picture
is something I will always aspire to.
%
You continually put your trust in me and found funding again and again so I
could continue my work.
%
Thank you for always believing in me, even in times when I doubted myself.
%

%
Dominique, thank you for opening up your group and resources to me as an
outsider and always making me feel a welcome and valuable part of your team.
%
Working at LSS has widened my horizon in so many ways, and I will always be
grateful for the experiences and opportunities it has given me.
%

%
Christian, maybe more than anyone else you have been a role model to me in the
last years.
%
We share an attention to detail and the desire to do things in the right way.
%
You are an amazing teacher and I have learned so much from you.
%
Keep doing what you are doing, I can envision nobody who can do it better.
%

%
Abou, it was a pleasure to be able to work with you.
%
Your dedication and work ethic amaze and inspire me.
%
Learning from each other with each discussion we had was a rich and fruitful
experience I will always remember.
%

%
Maik, although the year we spent working on \acs{FTLE} ridges together did not
amount to any publications and is completely absent from this dissertation,
I do not regret it at all.
%
I know few people whose passion for their work is as infectious as yours.
%
Working with you was probably the most fun I had doing research during my whole
time as a PhD student.
%

%
My co-authors Dirk, Tim, and Michael, thank you for your help and input which
contributed significantly to the success and quality of the contributions in
this thesis.
%

%
Stefan and Samuel, thank you for providing guidance on the subject of solid
mechanics, which I knew almost nothing about, and for finding the time to
produce data that I could use.
%
Without you, \cref{part:tensor_vis} of this thesis would probably be much
smaller and significantly weaker.
%

%
Petra and Petra, you are truly the rocks the ISG is built on.
%
Thank you for being there for everyone while making everyone feel like you are
there just for them.
%


%
Thoro, Heiko and Jürgen, thank you for keeping everything running smoothly so we
can all work in peace.
%


%
My office-mates and colleagues Thomas, Tim, Michael, Steve, Tobi, Janick, John,
Stefan, Anurag, Gordon, Felix, Cheng, G\'abor, and all the ones I have mentioned
before, you are truly amazing people.
%
Each of you is special to me in their own way.
%
I consider myself lucky to have been able to work with and be surrounded by such
an assortment of kind and interesting persons.
%

%
My parents Ruth and Michael, thank you for always supporting my curiosity and
for teaching me that if you decide to do something, you should do it right.
%

%
My wife Sarah, you are always there for me when I need you.
%
You cheer me up when I am feeling down and kick me when I become too complacent.
%
Thank you for placing the enormous amount of trust in me that you do.
%
I hope I deserve it.
%

%
My children Emil and Hannah, you give my life meaning in a way nothing and
nobody else does.
%
This thesis is just a side gig.
%
You are the real project.
%
I am convinced that by the time you can read this, you will be even more awesome
than you already are.
%

%
I also want to express my gratitude to the German state and the DFG, who funded
my education and research.
%
I am grateful to have grown up in a country that places such value in education
and research.
%

%
My final thanks goes to Filip Sadlo for agreeing to review this thesis.
%
May the following pages be worth your time.
%

%
Now, with the sentimental ramblings out of the way, let's get down to business.
%
% Holger
% Dominique
% Christian
% Abou
%
% Filip
%
% Gordon
% Felix
% Cheng
%
% Maik
% Dirk
% Tim
% Micha
%
% Stefan
% Samuel
% Philipp
%
% Thomas
% Steve
% Anurag
% John
% Janick
% Tobi
%
% Petra & Petra
%
% Mama & Papa
%
% Sarah, Emil & Hannah
%
% DFG, German State
%