
%
\section{Discussion} % (fold)
\label{sec:pev_discussion}
%
The \ac{PEV} operator was introduced as a generic operator, its interpretation
is dependent on the application scenario.
%

%
For stress tensors in mechanical engineering, the \ac{PEV} operator gives
insight into the alignment of the tensors under different acting forces.
%
Areas with \ac{PEV} lines can be wanted or unwanted.
%
In areas with present \ac{PEV} lines, the stress tensor is similarly oriented
for different external forces.
%
This could be used \eg, for deciding the placement of structural reinforcements
or to guide the selection of materials.
%
In regions without \ac{PEV} lines, there is no stress in a preferred direction
when applying different outer forces and material with a more isotropic behavior
could be used.
%

%
Besides this particular interpretation, there are general interpretations that
are common to all applications of the \ac{PEV} operator.
%
The \ac{PEV} operator is agnostic to isotropic scaling of the tensors.
%
It gives information about the orientation of the tensors only.
%
In this way, the \ac{PEV} operator can be seen as an addition to many standard
measures for comparing tensors like norm, trace, or eigenvalues.
%

%
The presented algorithm for piecewise linear tensor fields does not use any
derivatives of the data.
%
It depends on a number of thresholds to guide subdivision levels and filtering.
%
The spatial subdivision threshold $\varepsilon_\mathrm{d}$ influences the
accuracy of the resulting \ac{PEV} lines.
%
A small threshold means more subdivisions and is one of the main factors
influencing performance.
%
Since subdivision converges to single points, the computing time increases
logarithmically when decreasing $\varepsilon_\mathrm{d}$.
%

%
The directional subdivision threshold $\varepsilon_\mathrm{r}$ guides the
accuracy of the obtained eigenvector direction.
%
Typically, the smaller the current spatial triangle $\triangle_\vx$ becomes,
the smaller the region of valid eigenvector directions.
%
For increasing subdivision level in space, the valid eigenvector directions
will converge on a point.
%
This means that for small $\varepsilon_\mathrm{d}$, the recursion in the space
of directions will generally proceed to the highest subdivision level.
%
The influence of $\varepsilon_\mathrm{r}$ on computation time is about the same
as for $\varepsilon_\mathrm{d}$.
%
However, an accurate determination of eigenvector direction is essential to
decrease the number of candidate \ac{PEV} solutions that have to be clustered.
%
This means that $\varepsilon_\mathrm{r}$ should be chosen very small.
%
We find $\varepsilon_\mathrm{r} = \num{e-9}$ to be a choice that provides
consistently good results.
%

%
Because our algorithm can produce multiple candidate points for an intersection
of a \ac{PEV} line with a tetrahedral face, we need to cluster the results.
%
Theoretically, all candidate points should be in adjacent triangles, as
(unoriented) eigenvector directions in linear tensor fields do not oscillate
on small scales.
%
However, due to numerical noise and rounding errors on a computer, some
candidate triangles might not be exactly adjacent to each other.
%
To bridge this gap, the clustering threshold $\varepsilon_\mathrm{c}$ defines
the radius in which two candidate solutions are considered to belong to the same
cluster.
%
Because the numerical noise influencing the size of gaps between candidate
solutions is random, we do not expect candidates to be more than two or three
lengths of $\varepsilon_\mathrm{d}$ from each other.
%
We recommend to set $\varepsilon_\mathrm{c}$ to some fixed multiple of
$\varepsilon_\mathrm{d}$.
%
In our experiments, $\varepsilon_\mathrm{c} = 5 \varepsilon_\mathrm{d}$ proved
sufficient for all datasets.
%

%
The parallelism threshold $\varepsilon_\mathrm{p}$ is used to weed out false
positive candidates that are a byproduct of our algorithm.
%
It must be chosen carefully to separate false positive solutions from numerical
errors.
%
Because of this threshold, the spatial subdivision threshold
$\varepsilon_\mathrm{d}$ can not be chosen arbitrarily large.
%
The larger $\varepsilon_\mathrm{d}$, the larger the possible difference in
eigenvector direction between the real \ac{PEV} point and the tensor at the
center of the triangle, which is chosen as a representative.
%
In general, the choice of $\varepsilon_\mathrm{p}$ is dependent on
$\varepsilon_\mathrm{d}$.
%
More spatial subdivision levels enable a smaller choice of the parallelism
threshold.
%
In our experiments, a choice of $\varepsilon_\mathrm{p} = \num{e-2}$ for
$\varepsilon_\mathrm{d} = \num{e-3}$, and $\varepsilon_\mathrm{p} = \num{e-3}$
for $\varepsilon_\mathrm{d} = \num{e-6}$ worked very well.
%